Liveness-Based Garbage Collection

### 2014-09-10 paper/sections/liveness.tex
Just started

\section{The target language---syntax and semantics}
\label{sec:defs}
We let $x$, $y$, $z$ range over variables, $f$ over user-functions and $p$ over primitive
functions ($\CONS$, $\PRIM$ etc.).
The syntax  of our language is  shown in Figure~\ref{fig:lang-syntax};
it has eager semantics and restricts programs to be in 
Administrative  Normal Form (ANF)~\cite{chakravarty03perspective}
where all actual parameters to functions are variables.
This restriction does not affect expressibility (and indeed we feel free
to ignore it in examples when inessential), but simplifies the analysis formulation.
Additionally, as in the three-address instruction form familiar from compiler texts,
it forces each temporary to be named and function calls to be serialized
(necessary to get an unambiguous definition of liveness).
We further require that each variable in a program is distinct, so that no scope shadowing
occurs---this simplifies proofs of soundness.
In this formulation expressions: either perform a test ($\SIF$), make a computation step ($\LET$)
or return ($\RETURN$).
The $\RETURN$ keyword is logically redundant, but we find it clarifies the semantics and analysis.

The body of the program is
the  expression denoted  by  $e_\mainpgm$; for analysis purposes
it is convenient
to regard $e_\mainpgm$ as part of a function definition
$(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
% As  mentioned earlier,
We write $\pi\!:\!e$ to associate the  label $\pi$ (not part of the language syntax) with the program point just  before expression $e$.

% \paragraph{Inlining.}
In spite of the ANF restrictions it is still possible
to inline non-recursive functions (a fact we use to prove the safety
of liveness analysis).  A user-function call
$(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$
to a function defined
(after renaming its formals and locals to be disjoint from existing variables)
by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$
is replaced by a sequence of $\LET$'s of the form
$z_i\leftarrow (\ID\,\,y_i)$ followed by
the body $e_f$ but with its $(\RETURN\,\,w)$ expressions replaced by
$(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.

\begin{figure}[t!]
\footnotesize
\begin{eqnarray*}
   p \in \mathit{Prog} & ::= & d_1 \ldots d_n \,\, e_\mainpgm
    \hspace{7em} \mbox{\em --- program}\\
    d \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
    e) 
    \hspace{1.7em} \mbox{\em --- function definition} \\
e \in \mathit{Expr} & ::= &
\left\{\begin{array}{ll@{\hspace{4em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \mbox{\em --- conditional} \\ 
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) && \mbox{\em --- let binding} \\
       (\RETURN\,\, x) && \mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{Stmt} & ::= &
\left\{\begin{array}{lr@{\hspace{1em}}l}
       k && \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) && \mbox{\em --- constructor} \\ 
       (\CAR\,\, x) &  (\CDR\,\, x) & \mbox{\em --- selectors} \\ 
       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester and generic arithmetic} \\ 
%        (\NULLQ\,\, x) && \mbox{\em ---  tester} \\ 
       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)} \\ 
%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic} \\ 
       \multicolumn{2}{l}{(f\,\, x_1\,\,\ldots\,\, x_n)} 
            & \mbox{\em --- function application} 
    \end{array}\right.
\end{eqnarray*}
  \caption{The syntax of our language}\label{fig:lang-syntax}
\figrule
\normalsize
\end{figure}

\subsubsection{Semantics}
We now give an operational
semantics for  our language.  Later, a refinement of
the  operational semantics,  which we  call minefield  semantics, will
serve to prove liveness analysis correct.
We give a small-step semantics because, unlike big-step semantics,
correctness for non-terminating programs does not need special treatment.
We start with the domains used by the semantics:
\[
\begin{array}{rlcl@{\hspace{2em}}l}
% \begin{eqnarray*}
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Loc}& \mbox{-- Values}\\
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Val} &
\mbox{-- Environment} \\ 
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow ({Val\times
  Val}) & \mbox{-- Heap}
% \end{eqnarray*}
\end{array}
\]
Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
A value is either a number, the empty list $\NIL$, or a location  $\ell$.
Our liveness analysis does not track numeric values, and thus is neutral
as to whether these are boxed or represented as immediates.
An environment is a finite mapping from variables to values, and a heap a finite mapping
from locations to pairs of values.
% Given  a program,  an  environment is  a
% mapping from  the set  of variables of  the program  $\mathit{Var}$ to
% values in $\mathit{Val}$.    The heap is a mapping
% from  locations  to pairs of  values.
Finally, $\stk$ is a stack
(using $\bullet$ for push and $[\,]$ for empty stack)
of frames of unfinished function calls.  A
frame is a triple $(e,x,\rho)$ representing the call site
$
(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)
$
being evaluated in environment $\rho$.
Frames can also be viewed as continuations, in this view
the ({\sc ord-return}) 
 rule in the small-step operational semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.

The semantics 
of statements $s$ are given by the judgement form
$\rho, \heap, s \rightsquigarrow \heap', v$
and those for expressions $e$ by the form
$\rho, \stk, \heap, e \rightarrow \rho', \stk', \heap', e'$.
The start state is $(\{\},[\,], \{\}, e_\mainpgm)$ and the
program terminates successfully with result value $\rho(x)$
on reaching the halt state $(\rho,[\,], \heap, (\RETURN\ x))$

Notation:  we write $\rho[x \mapsto v]$ for the environment which is
as $\rho$ but has value $v$ at $x$.
We also write $[\myvec{x} \mapsto \myvec{v}]$ which respectively
has values $v_1, \ldots, v_n$ at $x_1, \ldots, x_n$ and write
$[\myvec{x} \mapsto \rho(\myvec{y})]$ when 
$v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.

\paragraph{Stuck states.} Note that certain forms of $e$ do not reduce
with $\rightarrow$ (perhaps because $\rightsquigarrow$ could not reduce
a contained $s$).  Some of these we eliminate syntactically, e.g.\
ensuring all variables and functions are defined and are called with
the correct number of parameters.
Others include $(\CDR\ \NIL), (\CAR\ 3), (+\ \NIL\ 4)$ and $(\SIF\ \NIL\ e_1\ e_2)$.
All but the first can be eliminated with a static type system but,
treating our program as dynamically typed, we regard all these
as stuck states.

% A program terminates successfully if it reaches the halt state
% in Figure~\ref{fig:lang-opsem-small}.
% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
% is the halt state (with result value $\rho(x)$).

\begin{figure}[t!]  
% (Random text to ensure proper line-spacing (problem with infrule macro?).)

\footnotesize
\begin{minipage}{0.335\textwidth}
\infrule[ord-const]
        {}
        {\rho, \heap, \pp{\pi}{k}
          \rightsquigarrow \heap, k}
\end{minipage}\begin{minipage}{0.655\textwidth}
\infrule[ord-cons]
      {
\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
}
      {\rho, \heap, \pp{\pi}{(\CONS\,\, x\ y)}
        \rightsquigarrow \heap[\ell\mapsto(\rho(x),\rho(y))],
        \ell}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
\infrule[ord-car]
        {\heap(\rho(x)) = (v_1, v_2)}
        {\rho, \heap, \pp{\pi}{(\CAR\,\, x)}
          \rightsquigarrow \heap, v_1}
\end{minipage}\begin{minipage}{0.595\textwidth}
\infrule[ord-cdr]
        {\heap(\rho(x)) = (v_1, v_2)}
        {\rho, \heap, \pp{\pi}{(\CDR\,\, x)}
          \rightsquigarrow \heap, v_2}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
\infrule[ord-id]
      {
}
      {\rho, \heap, \pp{\pi}{(\ID\,\, x)}
        \rightsquigarrow \heap, \rho(x)}
\end{minipage}\begin{minipage}{0.595\textwidth}
\infrule[ord-prim]
      {\rho(x) \in \mathbb{N} \andalso \rho(y)\in \mathbb{N}}
      {\rho, \heap, \pp{\pi}{(\PRIM\,\, x\ y)}
        \rightsquigarrow \heap, \rho(x) + \rho(y)}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
% \infrule[ord-null-false]
\infrule[]
        {\rho(x)\not= \NIL
        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
          \rightsquigarrow \heap, 0}
\end{minipage}\begin{minipage}{0.595\textwidth}
% \infrule[ord-null-true]
\infrule[ord-null]
        {\rho(x)= \NIL
        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
          \rightsquigarrow \heap, 1}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
% \infrule[ord-if-true]{
\infrule[]{
  \rho(x) \in \mathbb{N}\setminus\{0\}
}{
  \rho,\stk, \heap,
  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
  \longrightarrow \rho,\stk, \heap, e_1}
\end{minipage}\begin{minipage}{0.595\textwidth}
% \infrule[ord-if-false]{
\infrule[ord-if]{
  \rho(x) = 0
}{
  \rho,\stk, \heap,
  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
  \longrightarrow \rho,\stk, \heap, e_2}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-let-nonfn]{
%  \rho,\stk, \heap, \pp{\pi_1}{s} \longrightarrow \rho,\stk, \heap', v_1 
  \rho, \heap, \pp{\pi_1}{s} \rightsquigarrow \heap', v 
  \andalso
  \mbox{$s$ is not
$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
       }
}{
  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
    \pp{\pi_2}{e})}
  \longrightarrow
  \rho[x \mapsto v], \stk, \heap', \pp{\pi_2}{e}}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-let-fncall]{
  \mbox{$s$ is 
$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
       }
\andalso
\mbox{$f$ defined as $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \pp{\pi_f}{e_f})$}
}{
  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
    \pp{\pi_2}{e})}
  \longrightarrow
% [z_1\mapsto \rho(y_1), \ldots, z_n\mapsto \rho(y_n)], 
[\myvec{z} \mapsto \rho(\myvec{y})],
  \,(\rho,x,e)\bullet\stk,\, \heap, e_f
}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-return]{
}{
  \rho, \,(\rho',x',e')\bullet\stk,\, \heap,
  \pp{\pi}{(\RETURN\ x)}
  \longrightarrow \rho'[x' \mapsto \rho(x)], \stk, \heap, e'}
\end{minipage}

% \normalsize
% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
% is the halt state (with result value $\rho(x)$).

  \caption{The small-step operational semantics}
  \label{fig:lang-opsem-small}
\figrule
\normalsize
\end{figure}

%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point

@@

%% \comment{
%%   \begin{enumerate}
%% \item Set the framework for liveness. Basic notions: access
%%   paths and expressions.
%% \item An informal working definition of liveness. Minefield will
%%   provide the formal definition.
%% \item Liveness judgements. The form  of the judgement ($e, \sigma \len
%%   \Lv$)  and the  intuition  behind it.   Justification of  individual
%%   rules. One of the changes from  the earlier version is the lack of a
%%   post-liveness in  the judgements. This anticipates  the intended use
%%   of  the analysis  for garbage  collection and has  to be  carefully
%%   explained.
%%   \end{enumerate}
%% }

%% We now  set up  the framework for  liveness analysis.  Given  a memory
%% graph, a link in the graph  can be described by starting with a member
%% of  the  root   set  and  going  through  a   sequence  of  \CAR\  and
%% \CDR\ dereferences to reach the link.  For notational convenience, the
%% links corresponding to  the \CAR\ and \CDR\ fields  of the \CONS\ cell
%% are  denoted by  \acar\  and \acdr,  and  thus a  sequence of  several
%% dereferences    is    represented   by    a    string   $\alpha    \in
%% \lbrace\acar,\acdr\rbrace^*$.  The string  $x.\alpha$ describes a path
%% in the  memory graph that starts  at $x$ and  traverses adjacent links
%% specified by  $\alpha$.  $x.\epsilon$ is  written more simply  as $x$.
%% We call the string $x.\alpha$ an {\em access expression\/}, the string
%% $\alpha$ an {\em access pattern\/},  and the path traced in the memory
%% graph   by  the   access  expression   an  {\em   access   path}.   In
%% Figure~\ref{fig:mot-example},        the       access       expression
%% \ww.\acdr\acar\acar\ represents the access  path from \ww\ to the node
%% containing the value  $4$.  Most often the memory  graph is clear from
%% the context,  and we shall use  access expressions to  refer to access
%% paths. An  access path  used to refer  to a  link in the  memory graph
%% refers    to     the    last     link    in    the     path.     Thus,
%% \ww.\acdr\acar\acar\ denotes the link  incident on the node containing
%% the value $4$.  As a  notational convenience, if $\sigma$ represents a
%% set of access paths, then  $x.\sigma$ will represent the set of access
%% paths $  \lbrace x.\alpha \mid \alpha  \in \sigma\rbrace$.  Similarly,
%% if  $r$ is  the single  symbol \acar\  or \acdr,  then  $r\sigma$ will
%% denote the  set of access patterns  $ \lbrace r\alpha  \mid \alpha \in
%% \sigma\rbrace$.

%% \subsection{Liveness Analysis}
%% \label{sec:exp-liveness}
%% 
%% Consider the memory graph at a program point $\pi$ during an execution
%% of a program. A  link in the memory graph is said  to be {\em live} at
%% $\pi$,  if it  is used  (dereferenced or  tested) during  a  period of
%% execution  starting from  the  evaluation of  the expression  labelled
%% $\pi$, to the end of the program.  An access path $x.\alpha$ is said to
%% be live at  $\pi$, if, during {\em any} execution  of the program, the
%% link denoted by $x.\alpha$ in the memory graph is live at $\pi$.  Note
%% that an access path such as  $\px.\acdr\acar$ can be live in two ways:
%% In the first, the path  itself represents the sequence of dereferences
%% to  access  the last  edge,  as,  for  example, while  evaluating  the
%% expression  $(\LET\,\,   \pz  \leftarrow  (\CDR\,\,\,\px)\,\,  \IN\,\,
%% (\CAR\,\,\pz))$.  In the second, the  path may be made live indirectly
%% by  evaluating $(\LET\,\,  \pz \leftarrow  (\CAR\,\,\,\py)\,\, \IN\,\,
%% (\CAR\,\,\pz))$  in an execution  context where  $(\CDR\,\,\,\px)$ and
%% $(\CAR\,\,\,\py)$ share  the same node.   We call an access  path {\em
%%   explicitly  live},  if  it is  live  in  the  first sense  and  {\em
%%   implicitly  live} if it is live in the second sense.  Clearly, static
%% determination   of  implicit  liveness   requires  an   analysis  that
%% determines  potential sharing  during  any execution  of the  program.
%% However, the intended application  of our liveness analysis is garbage
%% collection  during which  we have  access  to the  entire heap.   This
%% obviates the need for a  separate sharing analysis and we restrict our
%% attention to explicit liveness only.
%% 
%% 
%% %% We  extend  our  alphabet   of  access  patterns  to  include  symbols
%% %% \bcar\ and \bcdr\   representing backward traversals over a  $\CAR\ $ or
%% %% $\CDR\  $  edge.   If  traversals  over \acar\  and  \acdr\  represent
%% %% dereferencing    or   selection,    traversals    over   \bcar\    and
%% %% \bcdr\ represent application of  the constructor \CONS.  The following
%% %% example motivates the need for these symbols.
%% 
%% %% \begin{example}\label{exmp:motiv-bar}
%% %%   Consider the expression at program point $\pi_1$ in
%% %%   \begin{center}
%% %%     {\sf $\pi_1\!:${(\LET\ w $\leftarrow$
%% %% 	                    $\pi_2\!:${(\CONS\ x y)} \scmin\ 
%% %% 			    $\pi_3\!:${$\cdots$})}}
%% %%   \end{center}
%% %%   Assuming  that  the   set  of  live  access  paths   at  $\pi_3$  is
%% %%   $\lbrace\ww.\alpha\rbrace$, we  would like to find  out which access
%% %%   paths  rooted   at  \xx\   are  live  at   $\pi_1$.   Let
%% %%   \xx.$\alpha'$ be live at $\pi_1$.  Then, the two possible cases are:
%% %%   \begin{itemize}
%% %%   \item If $\alpha = \acdr\beta$ or $\alpha = \epsilon$, then there is
%% %%     no link  that is accessible through  a path rooted at  \xx\ and is
%% %%     live at $\pi_1$.
%% %%   \item  If  $\alpha  =  \acar\beta$  then  the  link  represented  by
%% %%     $\ww.\alpha$  is also  represented  by $\xx.\beta\ $and  is live  at
%% %%     $\pi_1$.  Thus, $\alpha' = \beta$.
%% %%   \end{itemize}
%% %%   This  relation  between  $\alpha$  and  $\alpha'$  is  expressed  by
%% %%   $\alpha' = \bcar\alpha$.  \bcdr\ can be interpreted similarly. \qed
%% %% \end{example}
%% 
%% %% While the inclusion of \bcar\ and \bcdr\ in the alphabet enables us to
%% %% treat  applications  of  selectors  and  constructors  uniformly,  the
%% %% resulting access patterns may not  be directly useful in that they may
%% %% not describe a  path in the memory graph.  We  define a {\em canonical
%% %%   access pattern\/}  as a pattern  that is restricted to  the alphabet
%% %% $\lbrace\acar,  \acdr\rbrace$.   Given   an  access  pattern  $\alpha$
%% %% containing \bcar\ and \bcdr, it  may be possible in certain cases
%% %% to obtain a canonical access  pattern representing the same edge. This
%% %% is  through reductions  based on  the identities  $(\CAR\  (\CONS\ \px
%% %% \  \py))  =  \px$  and  $(\CDR\  (\CONS\  \px  \  \py))  =  \py$.   In
%% %% Figure~~\ref{fig:mot-example},   the  path  $\pz.\bcdr\acdr\acar\acar$
%% %% reduces  to $\pz.\acar\acar$  in this  sense.  If  $\alpha$  cannot be
%% %% reduced  to path without  \bcar\ and  \bcdr\ by  these rules,  then we
%% %% reduce it  to the special  canonical access pattern $\bot$.   Since an
%% %% access  path with  $\bot$ cannot  be used  for traversing  a  heap, it
%% %% cannot  be  used for  nullification  during  garbage collection.   The
%% %% liveness of a such a path is irrelevant, and the path is not considered
%% %% any further.
%% 
%% %% $\alpha  \rightk{k}  \alpha'$ denotes  the  reduction  of $\alpha$  to
%% %% $\alpha'$  in $k$ steps,  and $\rightstar$  denotes the  reflexive and
%% %% transitive closure of $\rightarrow$.  The concatenation ($\cdot$) of a
%% %% set of access patterns $\sigma_1$  with $\sigma_2$ is defined as a set
%% %% containing  concatenation  of each  element  in  $\sigma_1$ with  each
%% %% element in $\sigma_2$, i.e.
%% 
%% %% \begin{eqnarray*}
%% %%   \sigma_1\cdot\sigma_2 &\eqdef& \lbrace \alpha_1\alpha_2 \mid \alpha_1 \in
%% %%   \sigma_1, \alpha_2 \in \sigma_2 \rbrace
%% %% \end{eqnarray*}

@@

%%% save two lines to bring up section 3.2.

% The  next  subsection {\color{red}will  describe}  how  to  derive a  function  for
% \Lfonly\  from this  rule.  \Lfonly\  is a  summary of  functions that
% enables us to do a context-dependent analysis of a call without having
% to enter the body of the called function.

%% We present  our liveness  analysis using judgement forms $\len$
%% for both expressions and programs.
%% These are given formally in Figure~\ref{fig:live-judge}, but
%% first we an intuitive understanding of the issues involved.
%% Consider the expression:
%% 
%% \begin{center}
%% \begin{uprogram}
%% 	  \UFL\ \hspace*{-1\TAL} $\pi\!:$(\LET\, \pz\   $\leftarrow$ (\CAR\,\,
%%                         \px) \IN\  
%%            % \UNL{0}
%%              $\pi'\!:\!$ (\LET\, \py\  $\leftarrow$ (\PRIM\,  \pz\,\,
%%                         \pw ) \IN \ldots) 
%% \end{uprogram}
%% \end{center}
%% 
%% 
%% Assume as a starting point that liveness at $\pi'$ is $\{\pz.\epsilon,
%% \pw.\epsilon\}$.
%% Informally, this is so since \pw\ and \pz\ are used in
%% (\PRIM~\pz~\pw ).
%% 
%% \begin{enumerate}
%% \item The liveness  at $\pi$ is obtained by  removing the access paths
%%   starting  with \pz\  from the  liveness  at $\pi'$  and adding  the
%%   liveness  generated ($\mathit{ref}()$) by $(\CAR~\px)$\footnote{This is
%%     similar  to liveness  analysis of  imperative programs,  where the
%%     liveness before a statement $s$ is given as $\mathit{live_{in}}(s)
%%     =   \mathit{live_{out}}(s)    \setminus   \mathit{def\/}(s)   \cup
%%     \mathit{ref\/}(s)$. Here  $\mathit{live_{out}}(s)$ is the liveness
%%     after $s$ and  $\mathit{ref\/}(s)$ and $\mathit{def\/}(s)$ are the
%%     liveness generated and  killed in $s$}. 
%% \item  Since  \pz\ is  bound  to  the  value of  $(\CAR~\px)$,  the
%%   liveness of \pz, namely $\epsilon$, is transferred to $(\CAR~\px)$
%%   as a {\em  demand}.  The demand on an expression consists of  access
%%   patterns  representing all  possible future  uses of  its  value.  A
%%   demand  of $\epsilon$  means that  only the  reference resulting
%% %  from the evaluation  of
%%   from  $(\CAR\,\,\px)$ will be used.
%% \item  Because of this  demand, the  liveness generated  by (\CAR\,\,
%%   \px) are $\px$ and $\px.\acar$.  Thus liveness at $\pi$ is
%%   is $\{\px.\epsilon, \px.\acar, \pw.\epsilon\}$.
%% \end{enumerate}
%% 
%% In general,  an access path is  live before an  expression because the
%% link represented by it is used (i) while evaluating the expression, or
%% (ii) while using the result of the expression, or (iii) simply because
%% the link  is live at  a later program  point.  When the  expression in
%% consideration  is (\CAR~\px), (i)  is illustrated  by the  access path
%% \px, (ii)  is illustrated by the  access path \px.\acar,  and (iii) by
%% the access  path \pw.   Note that \px.\acar\  is live in  this example
%% because  the   result  of  (\CAR~\px)   is  used  in   the  expression
%% (\PRIM~\pz~\pw  ) through  the variable  \pz.  In  contrast, replacing
%% (\PRIM~\pz~\pw ) by the expression (\CONS~\pz~\pw) would not have made
%% (\CAR~\px) live, since \CONS\ would not have dereferenced \pz.
%%

% \paragraph{Judgements formally.}
% Let  $\Lv$  denote a  liveness  environment (a function from
% variables to access paths, abused using grammar notation) and
% $\sigma$ be a demand.
% The judgement form $e, \sigma \len \Lv$,
% reads  as:  if  the demand  on  $e$  is  $\sigma$, then  the  liveness
% environment  before $e$  is $\Lv$ and
% is thus analogous to the classical notion of {\em transfer function}.

%    It is  interesting to  note that,
% unlike traditional liveness analysis,  our judgements do not require a
% liveness after the expression. This  is because of the coming together
% of  two reasons.   First, it  will be  clear while  discussing garbage
% collection,  that it is  enough for  our purposes  to assume  that the
% post-liveness  of  each function  body  is  $\emptyset$.  Second,  the
% restriction of our  programs to ANF ensures that  the post-liveness of
% each  expression  $e$  inside   the  function  body  continues  to  be
% $\emptyset$. Thus the post-liveness can be dropped altogether.

@@

\normalsize
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%==============================================================

% The liveness judgement rules are shown in Figure~\ref{fig:live-judge}.
% Given  a  demand $\sigma$,  the  function  $\mathit{ref\/}$ gives  the
% liveness generated  before the right-hand-side of  a \LET\ definition.
% Constants    do   not   generate    any   liveness.     The   operator
% \PRIM\ dereferences its arguments  resulting in the liveness shown.  A
% demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
% $\acar\sigma$  on  $x$.   In  addition, \CAR\  also  dereferences  its
% argument.  This  generates the liveness  $\{x\} \cup  x.\acar\sigma$ before
% (\CAR~$x$).   In a  opposite  sense, the  demand  of $\acar\sigma$  on
% (\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
% that \CONS\ does not, by itself, dereference its arguments.
% 
% 
% Unlike primitives, the mapping from the demand on a user-function call
% to the  demands on its arguments  has to be derived  from the function
% body itself---as a fixed point for a recursive function.
% The function  \Lfonly\ captures this mapping for
% all functions  defined in the  program.  For a  function $\mathit{f}$,
% \Lf{f}{i}{\sigma} gives the transformation of the demand $\sigma$ from
% a    call     to    the     $i$th    argument.     The     rule
% (\rn{live-define})  says that  if  a  guess  of \Lfonly\  is
% confirmed by the body of all the functions, then the guess is correct.
% {\color{blue}Note the similarity of this rule to type-inference
% for recursive program schemes.}

% The   liveness   judgements    for   \RETURN\    and   \SIF\   are
% straightforward. $x.\epsilon$  appears in the liveness  of \SIF\ since
% $x$  is  tested as  a  conditional  variable.   Finally, as  mentioned
% earlier,  the liveness  before  a let  expression $(\LET~x  \leftarrow
% s~\IN~e)$ is  obtained by  first finding the  liveness before  the let
% body  $e$,  removing the  liveness  of $x$  from  it,  and adding  the
% liveness generated before $s$.

\subsection{Minefield semantics and correctness}
\label{sec:minefield}
\label{sec:correctness}

This section gives a modified semantics
%(a reference semantics)
which checks liveness annotations
at run time, and `explodes' when these are found to be inconsistent with execution behavior,
but otherwise behaves as the standard semantics.
We show that such explosions never occur and hence run-time checks
can be elided.
%% We first assume some demand transformers $\Lfonly$;
%% our proof will rely on this being a solution (i.e.\ satisfies $\mathit{Prog} \len \Lfonly$)
%% of liveness analysis, but we do rely in this yet.
We first assume an arbitrary  demand transformers $\Lfonly$ which satisfies
$\mathit{Prog} \len \Lfonly$.
We then enrich the abstract machine state $\rho, \stk, \heap, e$ to
$\rho, \stk, \heap, e, \sigma, \Sigma$.
Here $\sigma$ is the demand to the currently active function,
thus the liveness $\Lv$ at $e$ is $\mathcal{L}(e,\sigma,\Lfonly)$,\footnote{
This a simple liveness propagation using $\mathcal{L}(\cdots)$
and $\mathit{ref}(\cdots)$ as $\Lfonly$ is assumed given.}
and $\Sigma$ is a stack of demands---one for each function frame
pushed in $\stk$.

Second, we augment $\mathit{Val}$ with a value $\bot$.
To model strong liveness $\bot$ may be copied freely,
including into a $\CONS$ cell,
% but reduces the state to $\bang$
but explodes
when used computationally (in a primitive operation other than a copy).
Additionally we define
$\mathit{GC}(\Lv,\Sigma): (\rho, \heap, \stk) \mapsto (\rho', \heap', \stk')$
which determines {\em live-reachability}\footnote{Reachability curtailed by liveness information.}
%%%%% \marginpar{Say more about live-reachability?}
using $\rho$ and the $\rho$'s in $\stk$ as the root set
and following links in $\heap$ {\em only as far as allowed by $\Lv$ and $\Sigma$}.
% The effect of $\mathit{GC}(\Lv,\Sigma)$ is to replace
Hence $\mathit{GC}(\Lv,\Sigma)$ replaces
live-unreachable values---in $\rho$, in $\heap$ and the $\rho$'s in \stk---with $\bot$.
For example, if $x.\epsilon \not\in \Lv$ then $\rho'(x) = \bot$.
%, regardless of $\rho(x)$.
Only $\mathit{GC}(\cdots)$ introduces $\bot$.

Third, we update the semantics in four ways: ({\em i\/}) we arrange that all
($\rightarrow$) transitions on expressions $e$ first use $\mathit{GC}(\cdots)$
to update the state and then continue as before;
and ({\em ii\/}) whenever the value $\bot$ is used computationally in a reduction
($\rightsquigarrow$), we enter a distinguished stuck state $\bang$.
For example, supposing $\rho(x)=\bot$
then
$\rho, \heap, s \rightsquigarrow \bang$ if $s$ is
$(\CAR\,\, x)$, $(\CDR\,\, x)$ or $(\PRIM\,\, x\ y)$,
but not if $s$ is $(\ID~x)$, $(\CONS~x~y)$ or $(f~x~y)$.
Finally ({\em iii\/}) we make a similar change to the ($\rightarrow$) reduction for ({\sc ord-if})
and ({\em iv\/}) augment the ({\sc ord-let}) rule for primitives to propagate
$\bang$ from ($\rightsquigarrow$) to ($\rightarrow$).
%%%%%% {\color{blue}The full minefield semantics is given as an appendix?}

% Suppose liveness for the current $e$ under reduction is $\Lv$; then:
% For every variable $x$ for which $x.\epsilon \not\in \Lv$ then replace
% $\rho(x)$ with $\bot$.
% Also, for every $\ell \in \mathit{Loc}$, replace the $\CAR$ field
% of $\heap(\ell)$ with $\bot$ when
% which there is no path $x.\alpha\acar$ in $\Lv$ such that following
% $\alpha$-links in $\heap$ from $\rho(x)$ reaches $\ell$.
% Correspondingly, replace the $\CDR$ field of $\ell$ if there
% is no path $x.\alpha\acdr$ in $\Lv$.
% \item
% 
% {
% \footnotesize
% \begin{minipage}{0.445\textwidth}
% \infrule[mine-car]
%         {\rho(x) = \bot}
%         {\rho, \heap, {(\CAR\,\, x)}
%           \rightsquigarrow \bang}
% \end{minipage}
% \begin{minipage}{0.445\textwidth}
% \infrule[mine-cdr]
%         {\rho(x) = \bot}
%         {\rho, \heap, {(\CDR\,\, x)}
%           \rightsquigarrow \bang}
% \end{minipage}\\[1ex]
% 
% \begin{minipage}{0.445\textwidth}
% \infrule[mine-prim]
%       {\rho(x) = \bot \vee \rho(y) = \bot}
%       {\rho, \heap, {(\PRIM\,\, x\ y)}
%         \rightsquigarrow \bang}
% \end{minipage}
% \begin{minipage}{0.445\textwidth}
% \infrule[mine-null]
%         {\rho(x)= \bot
%         }{\rho, \heap, {(\NULLQ\,\, x)}
%           \rightsquigarrow \bang}
% \end{minipage}
% }
%

The resulting minefield semantics behaves identically
(identical heap, identical steps, including possible non-termination)
to the standard semantics, except for one possibility---the
minefield semantics going $\bang$ while the standard semantics
continues (either to a halt state, to a stuck state, or reduces forever).

>>>>

@@

@@

@@

\normalsize
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure}
